---
title: "robSynth: Robust Synthetic Data Generation"
author: "Matthias Templ"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
bibliography: refs.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{robSynth: Robust Synthetic Data Generation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  message = FALSE,
  warning = FALSE
)
```

## Abstract

**robSynth** generates synthetic microdata in the presence of contaminated
training data.  Standard sequential synthesisers such as
**synthpop** [@nowok2016] and **simPop** [@templ2017] use OLS or CART to
model each variable; when the training data contains outliers, coding
errors, or linkage mismatches, each conditional model is distorted and
the distortion compounds through subsequent variables.  **robSynth**
replaces these models with MM-estimation [@yohai1987; @maronna2019] for
continuous variables and Huber-weighted logistic regression for
categorical variables, and extends the toolbox with robust tree-based
methods (median-leaf CART, OOB-residual-weighted random forest,
pseudo-Huber XGBoost).  The package supports per-variable methods,
custom predictor matrices, survey weights, hierarchical (household)
data, multiple synthesis with Reiter's partial-synthetic combining
rules [@reiter2003], robust Box-Cox transformations [@boxcox1964;
@marazzi2006], utility diagnostics (pMSE, KS, regression preservation,
correlation-Frobenius distance) and disclosure-risk assessment (k-anonymity,
closest-record distance, TCAP, and RAPID [@thees2025]).  This vignette
walks a statistical officer or applied methodologist through the package
end-to-end.

## 1. Introduction

Synthetic microdata are an increasingly standard release strategy for
National Statistical Offices and research data centres: a generative
model is fit to the real data, and the released artefact is a draw
from that model rather than the records themselves.  Under the
conditional synthesis framework that underpins **synthpop** and
**simPop**, each variable $V_j$ is drawn from a model
$V_j \mid V_{<j}$, so any bias in the conditional model directly
contaminates the synthetic output.

Real surveys contain messy data: typos, sentinel codes that leak
unreplaced into a numeric column, linkage mismatches, and genuine
extreme observations.  OLS and CART have zero breakdown point — a
single pathological row can drive the synthesiser arbitrarily far from
the majority.  **robSynth**'s contribution is to replace each
conditional model with a bounded-influence estimator.  The guarantee
is qualitative: under a contaminated Gaussian model, the error of the
OLS-based synthesiser grows without bound in the contamination
magnitude $\delta$, while the MM-based synthesiser's error is bounded
by a constant.

The remainder of this vignette is organised as follows.  Section 2
describes the methods and the key formulae.  Sections 3 and 4 cover
installation and a worked contamination example.  Sections 5–11 walk
through practitioner-oriented features (output structure, method
choice, per-variable methods, transformations, formulas, predictor
matrix, sentinel-coded missings).  Sections 12–19 cover inference and
diagnostics (multiple synthesis, utility, comparison, disclosure risk,
hierarchical data, survey weights, parallelisation).  Sections 20–21
discuss fit-failure diagnosis and migration from **synthpop**.
Section 22 summarises.

## 2. Methods

### 2.1 Conditional synthesis framework

Given data $X$ with variables $V_1,\ldots,V_p$ in visit order, **robSynth**
draws each synthetic variable from a fitted conditional model:
$$
\tilde V_j \sim \hat f_j\bigl( \cdot \mid \tilde V_1, \ldots, \tilde V_{j-1} \bigr),
$$
where $\hat f_j$ is trained on the **original** predictors $V_{<j}$
(not the already-synthesised ones).  A *predictor matrix* $P \in \{0,1\}^{p \times p}$
lets the user zero out specific conditioning variables; *custom
formulae* allow interactions and passive relationships.

### 2.2 MM-estimation for continuous variables

OLS minimises $\sum(y_i - x_i^\top\beta)^2$, so a single gross outlier
contributes arbitrarily much to the objective — its leverage can drag
the fitted slope to any value.  MM-estimation [@yohai1987] replaces
the squared error with a **bounded** loss $\rho$: once a residual
exceeds a tuning threshold, its contribution stops growing, and no
single row can dominate.  Concretely, the default continuous method
(`method = "mm"`) fits each conditional model by
$$
\hat\beta \;=\; \arg\min_{\beta}\; \sum_{i=1}^{n} \rho\!\left(\frac{y_i - x_i^\top\beta}{\hat\sigma}\right),
$$
with $\rho$ the Tukey bisquare function and $\hat\sigma$ an S-estimate
of scale.  Under Gaussian errors this achieves 50 % breakdown point
— the maximum possible fraction of arbitrary contamination the
estimator tolerates before the estimate can be driven to infinity —
and, with appropriate tuning, 95 % asymptotic efficiency
[@maronna2019].

### 2.3 Robust logistic regression for categorical variables

For binary targets (`method = "robust_logreg"`), the default fits a
GLM, computes standardised Pearson residuals $r_i^*$, and refits with
observation weights $w_i = 0.1$ where $|r_i^*| > 3$ and $w_i = 1$
otherwise.  For multinomial targets, the same downweighting is applied
to a prior-standardised log-score $\log p_i - \log \pi_{y_i}$, so that
rare minority-class rows are not blanket-downweighted.

### 2.4 Robust tree-based methods

- `robust_cart`: rpart-fitted tree, but leaf predictions are the
  weighted median of in-bag response values rather than the mean.  Leaf
  lookup at predict time uses `partykit::as.party()` node IDs, so fit
  and predict stay in sync even when several leaves share a mean.
- `robust_rf`: two-pass.  An initial forest produces out-of-bag
  residuals; rows with residual z-score $> 3$ are downweighted and the
  forest is refit.  Because the flag uses *conditional* residuals,
  legitimate extreme-$y$ rows driven by $X$ are not penalised.
- `robust_xgboost`: XGBoost with `reg:pseudohubererror` loss; training
  factor levels are stored via `xlev` and re-used at predict time to
  prevent silent column-mismatch bugs when new data lacks a rare level.

### 2.5 Robust Box-Cox

For skewed variables the package offers `transform = list(var = "boxcox")`
with a robust lambda search: the minimum of
$|\mathrm{mean}(y^{(\lambda)}) - \mathrm{median}(y^{(\lambda)})|\,/\,\mathrm{MAD}(y^{(\lambda)})$
over a grid.  This is itself resistant to the outliers that motivated
transforming in the first place [@marazzi2006].

### 2.6 Multiple synthesis: Reiter's combining rules

When $m > 1$ synthetic copies are generated, `synth_lm()` and
`synth_glm()` fit the analysis model on each copy and pool the
estimates.  Why not simply reuse Rubin's multiple-imputation rules?

In classical multiple imputation the analyst's uncertainty has **two**
sources: (i) sampling the population and (ii) imputing within the
sample.  Rubin's total variance is therefore $\bar U + (1 + 1/m)\,B$,
where $B$ accounts for imputation variability and the $(1+1/m)$
accounts for the sampling step having been performed *once*.  Synthetic
data generated by `robsynth()` has **only** source (ii) — the analyst
already conditions on the original sample, so there is no additional
sampling uncertainty to absorb.  Applying Rubin's rules therefore
inflates the variance by a factor proportional to $(1+1/m)$, producing
intervals that are anti-conservative in the reverse direction: wider
than they should be.

Reiter [@reiter2003] replaces $(1+1/m)B$ by $B/m$, scaling the
between-copy variance down because it reflects only synthesis noise:
$$
\bar Q = \frac{1}{m}\sum_{k=1}^m \hat Q^{(k)}, \quad
\bar U = \frac{1}{m}\sum_{k=1}^m \hat U^{(k)}, \quad
B = \frac{1}{m-1}\sum_{k=1}^m (\hat Q^{(k)} - \bar Q)(\hat Q^{(k)} - \bar Q)^\top,
$$
$$
\boxed{\,T_p \;=\; \bar U \;+\; B / m\,}, \qquad
\nu = (m-1)\bigl(1 + m\,\bar U / B\bigr)^2.
$$
The degrees-of-freedom formula $\nu$ inflates when $\bar U \gg B/m$
(synthesis noise small relative to within-copy variance) and shrinks
when $B/m \gg \bar U$ (synthesis adds substantial noise).  Single-copy
synthesis ($m=1$) sets $B=0$ exactly and recovers the ordinary
single-dataset variance with $\nu = \infty$.

## 3. Installation

**robSynth** is distributed from GitHub.  A CRAN release is planned
once the companion paper is accepted.

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("matthias-da/robSynth")
```

Several methods rely on optional backends listed under **Suggests** in
the DESCRIPTION: `partykit` (for `ctree` and for `robust_cart`
prediction), `ranger` (for `robust_rf`), `xgboost` (for
`robust_xgboost`), and `riskutility` (for mixed-type disclosure
measures).  Install them on demand:

```{r install_suggests, eval = FALSE}
install.packages(c("partykit", "ranger", "xgboost", "riskutility"))
```

The remainder of the vignette assumes you have run
`library(robSynth)` in your R session — the code chunks do so at the
start of the worked example.  Every subsequent example builds on the
data frame `df` created in Section 4, so it is best to read the
sections in order on first pass.

## 4. A worked example: robust vs non-robust under contamination

This is the flagship demonstration.  We generate a clean dataset with
a known linear structure, deliberately contaminate 15 % of the rows
with a large outlier shift, and then run two synthesisers side by
side: `robsynth()` with its robust default (MM for continuous,
robust-weighted logistic for categorical) and with
`method = "parametric"` (OLS for continuous, plain logistic for
categorical — the classical **synthpop**-style recipe).

The comparison is made against a known ground truth — the
regression coefficients of $X_4$ on $(X_1, X_2, X_3)$ — which is
fair because we built the data to have that structure.  In a real
application you would not have a ground truth; Section 13 discusses
the "contamination paradox" you face when comparing against the
contaminated training data alone.

### 4.1 Generating the clean data

We build a small synthetic population of 500 records with four
continuous variables and two binary covariates.  Unlike a typical
"independent normals" textbook example, $X_1, X_2, X_3$ here are
drawn jointly from a correlation structure that looks like something
you might see in a survey — $X_1, X_2$ are strongly correlated ($\rho
= 0.5$), $X_2, X_3$ moderately ($\rho = 0.4$), $X_1, X_3$ mildly
($\rho = 0.3$).  $X_4$ is a linear function of the three continuous
predictors with additive Gaussian noise; the factors $G_1, G_2$ are
logistic in $X_1$ and $X_2$ respectively, so they are also correlated
with the continuous variables.  This is enough structure to make
Figures 1–3 in Section 15 tell a non-trivial story.

```{r demo_build}
library(robSynth)

set.seed(42)
n <- 500

# Correlation structure for (X1, X2, X3)
Sigma <- matrix(c(1.0, 0.5, 0.3,
                  0.5, 1.0, 0.4,
                  0.3, 0.4, 1.0), nrow = 3, byrow = TRUE)
X123  <- MASS::mvrnorm(n, mu = c(0, 0, 0), Sigma = Sigma)
df    <- data.frame(X1 = X123[, 1], X2 = X123[, 2], X3 = X123[, 3])

# True relationship: X4 = 0.5*X1 + 0.5*X2 + 0.3*X3 + noise
df$X4 <- 0.5 * df$X1 + 0.5 * df$X2 + 0.3 * df$X3 + rnorm(n, sd = 0.5)
df$G1 <- factor(rbinom(n, 1, plogis(0.3 * df$X1)))
df$G2 <- factor(rbinom(n, 1, plogis(0.3 * df$X2)))
```

### 4.2 Contaminating 15 % of the rows

We store the clean data in `df_clean` **before** injecting
contamination, so we can later measure utility against the truth.
The contamination itself is a *block-shift* scenario: the same 75
rows (15 % of $n = 500$) get $+8$ added to $X_1$ and $X_2$
simultaneously, and their $G_1$ label is flipped.  This is a fairly
severe attack — eight standard deviations is far from typical "data
entry" error, but it makes the robust-vs-non-robust contrast visually
obvious.  Section 4a below repeats the exercise with a gentler
cellwise pattern that is closer to real surveys.

Note the factor-flip idiom: we capture the original two-level set
with `levels(df$G1)` and re-factor with the same `levels = ...` to
avoid silently re-coding the factor.  This is the kind of subtle bug
that contamination-injection code tends to introduce.

```{r demo_contaminate}
# Store clean version BEFORE contaminating (so factor levels are preserved)
df_clean <- df

# 15 % contamination: +8 on X1 and X2; flip G1 on the same rows
idx <- sample(n, 75)
df$X1[idx] <- df$X1[idx] + 8
df$X2[idx] <- df$X2[idx] + 8
g1_lev  <- levels(df$G1)
flipped <- ifelse(df$G1[idx] == g1_lev[1L], g1_lev[2L], g1_lev[1L])
df$G1[idx] <- factor(flipped, levels = g1_lev)
```

### 4.3 Robust vs non-robust synthesis, averaged over 5 replications

We now generate five synthetic copies with each method, fit the same
regression on each copy, and record two utility numbers per
replication: the root-mean-square error of the recovered regression
coefficients against the known truth
(`RMSE-beta`), and the Frobenius norm of the gap between the
original-clean and synthetic correlation matrices
(`Correlation-Frob.`).  Five replications is enough to see the
pattern without making the vignette slow to knit; a real benchmark
would use hundreds.

```{r demo_run}
# 5 replications (keeps the vignette knit under ~10 s)
true_coef <- c(X1 = 0.5, X2 = 0.5, X3 = 0.3)
R <- 5
rmse_mm <- rmse_ols <- cor_mm <- cor_ols <- numeric(R)
for (r in seq_len(R)) {
  s_mm  <- robsynth(df, seed = r)$synth                             # robust default
  s_ols <- robsynth(df, method = "parametric", seed = r)$synth      # OLS / plain logistic
  fit_mm  <- coef(lm(X4 ~ X1 + X2 + X3, data = s_mm))[c("X1","X2","X3")]
  fit_ols <- coef(lm(X4 ~ X1 + X2 + X3, data = s_ols))[c("X1","X2","X3")]
  rmse_mm[r]  <- sqrt(mean((fit_mm  - true_coef)^2))
  rmse_ols[r] <- sqrt(mean((fit_ols - true_coef)^2))
  cor_mm[r]   <- sqrt(sum((cor(df_clean[, 1:4]) - cor(s_mm[, 1:4]))^2))
  cor_ols[r]  <- sqrt(sum((cor(df_clean[, 1:4]) - cor(s_ols[, 1:4]))^2))
}

tab <- data.frame(
  Method              = c("MM (robust)", "OLS (parametric)"),
  `RMSE-beta`         = c(mean(rmse_mm), mean(rmse_ols)),
  `Correlation-Frob.` = c(mean(cor_mm),  mean(cor_ols)),
  check.names = FALSE
)
knitr::kable(tab, digits = 3,
             caption = "Table 1: Mean utility over 5 synthetic replications.")
```

**Reading the table.**  On RMSE-$\beta$, MM preserves the true
regression coefficients of $X_4$ on $\{X_1, X_2, X_3\}$ much better
than OLS: the contaminated rows pull OLS slopes toward the shifted
cloud (this is the classical *attenuation bias* — the least-squares
fit tries to accommodate both the bulk and the shifted block at
once), while MM's bounded $\rho$-function limits each contaminated
row's contribution to the objective.  On the correlation-Frobenius
distance the ranking is more muted, because this metric averages
over all 16 pairwise entries of the $4 \times 4$ correlation matrix,
and only the $(X_1, X_2)$-related entries are affected.  **Take-away:
the right utility metric depends on what you will do with the
synthetic data.**  If the downstream analysis is a regression,
RMSE-$\beta$ is the right target; if it is a correlation-based
exploratory analysis, the Frobenius gap is.  Section 22 returns to
the population-level story; the single-replication ranking here is
not a population claim.

### 4a. Cellwise scenario: response-side outliers

The contamination in §4.2 was *block*: the same 15 % of rows were
shifted on $X_1$ and $X_2$ simultaneously.  Block contamination is
easy to catch because whole rows misbehave — any row-level
Mahalanobis test will flag them.  In real data, however,
contamination is typically *cellwise* [@raymaekers2024]: a data-entry
typo in $X_2$ on row 42, a measurement glitch in $X_1$ on row 103, a
sentinel code leaked into $X_4$ on row 350.  The bad cells are
mostly on *different* rows.

For this section we focus on the cleanest cellwise case: **10 % of
response-side cells are shifted** (here, cells of $X_4$ which is the
response in the evaluation regression).  A reasoning note on why
this particular setup matters is given below the code.

```{r cellwise}
set.seed(11)
df_cw <- df_clean
cell_prob <- 0.10
flip <- runif(nrow(df_cw)) < cell_prob
df_cw$X4[flip] <- df_cw$X4[flip] + 15   # response-side outliers

R_cw <- 5
rmse_mm_cw <- rmse_ols_cw <- numeric(R_cw)
for (r in seq_len(R_cw)) {
  s_mm_cw  <- robsynth(df_cw, seed = r)$synth
  s_ols_cw <- robsynth(df_cw, method = "parametric", seed = r)$synth
  rmse_mm_cw[r]  <- sqrt(mean((coef(lm(X4 ~ X1 + X2 + X3, data = s_mm_cw ))[2:4]
                               - true_coef)^2))
  rmse_ols_cw[r] <- sqrt(mean((coef(lm(X4 ~ X1 + X2 + X3, data = s_ols_cw))[2:4]
                               - true_coef)^2))
}

tab_cw <- data.frame(
  Method      = c("MM (robust)", "OLS (parametric)"),
  `RMSE-beta` = c(mean(rmse_mm_cw), mean(rmse_ols_cw)),
  check.names = FALSE
)
knitr::kable(tab_cw, digits = 3,
             caption = "Table 1b: Cellwise response contamination (10 %, shift = 15; mean over 5 replications).")
```

**Why this demo is cleaner than predictor-side cellwise.**
`robSynth()`'s sequential recipe fits MM on each conditional model
$V_j \mid V_{<j}$.  MM bounds the influence of $y$-outliers, so when
$X_4$ (the response in its own conditional regression) has
cellwise-shifted values, MM downweights them and the fitted slopes
stay near the truth.  The downstream evaluation regression
`lm(X4 ~ X1 + X2 + X3, data = synth)` then recovers the true
coefficients, and OLS synthesis — which was *not* robust at the
fitting step — does not.

**Why predictor-side cellwise is harder.**  Contamination in the
*predictor* $X_1$ is a different problem.  The first variable in
the visit sequence is drawn *marginally* from the empirical
distribution (a weighted resample + kernel bandwidth), and the
marginal draw is **not** robust to cellwise outliers in that
variable — the 10 % of shifted $X_1$ cells get sampled into the
synthetic $X_1$.  MM on subsequent conditional models handles the
response side, but the contamination in $X_1$ remains and
propagates.  If you want full cellwise protection across all
variables, use `method = "robust_rf"` or `method = "robust_xgboost"`
(which downweight outliers inside tree fits) or pre-process the
original data with a cellwise filter (e.g. DDC, cellMCD) before
calling `robsynth()`.  The companion paper
[Templ 2026, submitted] discusses these trade-offs; full cellwise
preprocessing integrated into `robsynth()` is planned for a
future release.

## 5. What `robsynth()` returns

Calling `robsynth(df)` returns a list-of-class `"robsynth"` bundling
the synthetic data together with the configuration and the fitted
conditional models.  The `print()` method summarises the object
without dumping the full synthetic data frame.

```{r basic}
res <- robsynth(df, seed = 1)
print(res)
```

For almost every downstream use you only need `$synth`.  The other
components are kept for auditing, reproducibility, and diagnostics:

- `$synth` — the synthetic data frame (a *list* of data frames when
  `m > 1`).  Factor levels match the original, NAs from `cont.na` are
  propagated, and you can write it straight to CSV with
  `write.csv(res$synth, ...)`.
- `$original` — the training data stored verbatim.  Convenient because
  diagnostic functions like `utility(res)` pull it automatically;
  **remove with `strip_diagnostics(res)` before releasing** the
  synthetic copy to external users, or you would effectively be
  releasing the training data along with it.
- `$m`, `$method`, `$visit.sequence`, `$predictor.matrix`,
  `$transform_info` — the configuration used for the synthesis,
  exactly as resolved at call time.  Useful as a self-documenting
  audit trail.
- `$models` — the fitted conditional models from the last synthetic
  copy.  `str(res$models, max.level = 1)` gives a quick overview.
  Inspect these when a synthetic column looks suspicious (Section 20
  on convergence failures); do **not** release them because they
  encode information about the original data.
- `$var_types` — named character vector with `"continuous"` or
  `"categorical"` per column.  Useful for downstream scripts that
  need to handle the two types differently.

In practice the idiom is:

```{r basic_usage, eval = FALSE}
res <- robsynth(df, seed = 1)
write.csv(res$synth, "synthetic_release.csv", row.names = FALSE)
save(strip_diagnostics(res), file = "synthetic_release.rda")
```

## 6. Choosing a method

The defaults (`mm` for continuous, `robust_logreg` for categorical)
work well for the typical "messy official-statistics" case; the other
ten methods exist for specific problems.  Use the following rules of
thumb as a starting point and refine based on `utility()` diagnostics:

- **Clean data, linear structure** → `norm` (parametric, fastest) or
  `cart` (captures non-linearities).  If you are confident there is
  no contamination, there is no reason to pay the overhead of robust
  estimators.
- **Contaminated continuous with linear structure** → `mm` (the
  default).  This is the sweet spot: the linear form is correctly
  specified and the robustness buys you protection against the
  outliers.
- **Contaminated continuous, non-linear** → `robust_rf`
  (two-pass random forest with OOB-residual weights) or
  `robust_xgboost` (pseudo-Huber loss).  Use `robust_rf` when
  interpretability matters, `robust_xgboost` when you need to push
  predictive performance and are willing to pay the fit cost.
- **Heavy-tailed or skewed continuous** → `mm` combined with
  `transform = list(var = "boxcox")` (Section 8).  The robust
  Box-Cox lambda estimator (`robust_boxcox_lambda`) is itself resistant
  to the outliers that motivated the transformation.
- **Binary or multinomial with rare classes** → `robust_logreg`
  (default).  For clean multi-level data without concern for
  contamination, `polyreg` is faster.
- **Strictly derived variable** (e.g. $BMI = \text{weight} /
  \text{height}^2$, or a total income that is the sum of components)
  → `passive` with a `formulas` entry (Section 9).  Never synthesise
  a derived quantity directly; synthesise its components and
  re-derive.
- **Variables that must be kept verbatim** (e.g. geographic region,
  year-of-birth for epidemiology cohorts) → `sample` (marginal
  resampling) on those variables; leaves the variable identical to
  a bootstrap of the original values.

When in doubt, run `robsynth()` with the default and inspect
`utility(res)` against clean data (Section 13).  If the pMSE is much
larger than the theoretical null $c(1-c)/n$, try `robust_rf` or
`robust_xgboost` — a large pMSE typically flags a missed
non-linearity rather than a contamination issue.

You can also mix methods across variables (Section 7), which is the
right move whenever different columns have different problems.

## 7. Per-variable methods

The `method` argument accepts either a single string (applied to all
variables of the appropriate type) or a **named character vector**
whose names match the columns of `data`.  The example below pairs
different backends to different columns: `X1` gets plain MM, `X2`
gets the two-pass random forest because we suspect a non-linear
conditional mean, `X3` gets gradient boosting, and the factors get
logistic regression — robust on $G_1$ (we flipped it earlier, so it
is contaminated) and plain on $G_2$ (untouched).

Anything you leave unnamed gets the corresponding default:
continuous columns fall back to `mm`, categorical columns to
`robust_logreg`.  That lets you override only the variables you have
an opinion about.

```{r pervar}
res2 <- robsynth(df,
  method = c(X1 = "mm", X2 = "robust_rf", X3 = "robust_xgboost",
             X4 = "mm", G1 = "robust_logreg", G2 = "logreg"),
  seed = 1)
res2$method
```

The table below summarises the backends and their optional package
dependencies.  The robust variants in **bold** are the ones you will
typically want under contamination; the non-robust classics are kept
for compatibility with `synthpop` / `simPop` pipelines.

| Method | Type | Robustness | Requires |
|--------|------|-----------|----------|
| `"mm"` | Continuous (default) | MM-estimation, 50 % breakdown | — |
| `"robust_cart"` | Both | Median leaves (partykit node IDs) + MAD scale | **partykit** |
| `"robust_rf"` | Both | Case-weighted by OOB-residual magnitude (y-outliers conditional on X) | **ranger** |
| `"robust_xgboost"` | Both | Pseudo-Huber loss, xlev-aligned predict | **xgboost** |
| `"robust_logreg"` | Categorical (default) | Standardised-Pearson downweighting | — |
| `"norm"` | Continuous | Standard OLS (no robustness) | — |
| `"cart"` | Both | Standard CART | — |
| `"ctree"` | Both | Conditional-inference tree | **partykit** |
| `"logreg"` / `"polyreg"` | Categorical | Standard logistic / multinomial | — |
| `"sample"` | Both | Marginal resampling | — |
| `"passive"` | Derived | Deterministic formula | — |

Packages listed under *Requires* are in **Suggests**; install them
on demand for the corresponding method.  If the package is missing,
`robsynth()` errors with an informative message rather than silently
falling back.

## 8. Transformations for skewed variables

Many real variables (income, reaction times, counts, distances) are
heavily skewed.  MM-estimation assumes approximately symmetric
conditional errors, so fitting `mm` directly on an income variable
will systematically mis-fit the right tail.  The standard remedy is
a Box-Cox transform [@boxcox1964] before synthesis and an inverse
transform on the synthetic output.

The classical Box-Cox lambda is estimated by maximum likelihood,
which is itself sensitive to outliers — exactly the situation we are
trying to robustify against.  `robust_boxcox_lambda()` minimises a
robust skewness proxy, $|\text{mean}-\text{median}|/\text{MAD}$, over a
user-specified grid [@marazzi2006].

In the example below we add an income-like variable via `rlnorm()`,
inflate the already-contaminated rows tenfold to create a heavy tail,
and let `transform = list(income = "boxcox")` pick the lambda
automatically.  `lambda` near 0 corresponds to the log transform,
which is the expected answer for a log-normal input.

```{r transform}
set.seed(123)
df$income <- rlnorm(n, 10, 0.5)
df$income[idx] <- df$income[idx] * 10                      # contaminate

# Robust Box-Cox normalises before synthesis, back-transforms after
res3 <- robsynth(df, transform = list(income = "boxcox"), seed = 1)
cat("Estimated lambda:", res3$transform_info$income$lambda, "\n")
# Lambda near 0 = log-transform (expected for log-normal data)
```

Supported transforms are `"log"` and `"sqrt"` (fixed), and
`"boxcox"` (estimated).  Each is applied to the **original** column
before fitting any conditional model, and the inverse is applied
after synthesis — so the synthetic income column is on the original
scale.  The shift used to guarantee strict positivity is recorded in
`res3$transform_info` so downstream consumers can reverse the
transform manually if needed.

## 9. Custom formulas and visit sequence

`visit.sequence` controls the order in which variables are synthesised.
Each variable is conditioned on the variables that come *before* it
in the sequence.  The default visit order matches the order of columns
in `data`, which places categorical covariates *after* the continuous
variables — sometimes you want the opposite (e.g. synthesise
demographic factors first so that continuous variables can condition
on them).

`formulas` overrides the default "right-hand side = all prior
variables" rule for individual variables.  In the example below, the
`X3 = ~ X1 * X2 + G1` entry tells the synthesiser to condition $X_3$
on the interaction between $X_1$ and $X_2$ plus the factor $G_1$,
instead of the default additive conditioning on all prior variables.
This is the cleanest way to inject interactions, polynomials, or
passive-style relationships.

```{r formulas}
res4 <- robsynth(df,
  visit.sequence = c("G1", "G2", "X1", "X2", "X3", "X4", "income"),
  formulas = list(X3 = ~ X1 * X2 + G1),
  transform = list(income = "log"),
  seed = 1)
```

Only variables mentioned in `formulas` get a custom formula; every
other variable keeps the default.  If your analysis model has many
interactions, specify them here rather than trying to encode them
via the predictor matrix in Section 10 — a formula is expressive
where a matrix is binary.

## 10. Custom predictor matrix

The **predictor matrix** is the familiar **synthpop** / **mice**
idiom for specifying which variables condition which.  Row $i$ of
the matrix is the *in-list* for variable $V_i$: a $1$ in column $j$
means "include $V_j$ as a predictor when synthesising $V_i$".  By
default the matrix is lower-triangular with respect to
`visit.sequence`, so each variable conditions on all its
predecessors.

Why override the default?  Three common reasons:

- **Prevent leakage of a sensitive conditioning variable** (e.g. you
  want income synthesised *without* conditioning on geographic
  region, to break a disclosure path).
- **Avoid multicollinearity** when two predictors are nearly
  identical.
- **Enforce domain assumptions** (e.g. "BMI is only allowed to
  depend on weight and height, not on education").

In the example below we keep the default lower-triangular
conditioning for every variable except `X4`, which we restrict to
depend on $(X_1, X_2, X_3)$ only — dropping `G1` and `G2` which
the default would otherwise include.

```{r predmat}
p  <- ncol(df)
pm <- matrix(0L, p, p, dimnames = list(names(df), names(df)))
pm[lower.tri(pm)] <- 1L                            # default lower-triangular
pm["X4", ] <- 0L                                   # override X4's row:
pm["X4", c("X1", "X2", "X3")] <- 1L                #   condition on X1, X2, X3 only
res_pm <- robsynth(df, predictor.matrix = pm, seed = 1)
```

Two quick gotchas:

1. If row $i$ is all zeros, $V_i$ is synthesised **marginally**
   (no conditioning) using the `"sample"` fallback — this is
   sometimes useful deliberately but usually indicates an oversight.
2. The matrix must be compatible with the visit sequence: you
   cannot condition $V_i$ on $V_j$ if $V_j$ is visited *after*
   $V_i$.  `robsynth()` does not reorder for you.

## 11. Sentinel-coded missing values (`cont.na`)

Many NSO datasets encode "refused", "don't know", or "not applicable"
as numeric sentinels (`-9`, `-999`, `9999`, `.A`, `.B`, ...) rather
than `NA`.  If you feed such values to a synthesiser without
recoding, two bad things happen:

1. Variable-type detection may flag the column as "categorical"
   because of the concentration of sentinel values, or estimate
   its MAD from a distribution that includes `-999` points, turning
   the column's scale estimate into nonsense.
2. The sentinel value reappears in the synthetic output — and
   downstream users assume it is a valid observation.

The `cont.na` argument recodes sentinel values to `NA` **before**
type detection and **before** any scale estimator is invoked, which
is the only timing that lets the rest of the pipeline treat the
sentinels correctly.  It is also equivalent to the `cont.na`
argument in **synthpop**.

```{r cont_na}
df_na <- df
df_na$income[1:20] <- -9                                   # "refused"
res_na <- robsynth(df_na, cont.na = list(income = -9), seed = 1)
# The sentinel does not reappear in the synthetic output
any(res_na$synth$income == -9, na.rm = TRUE)
```

`cont.na` accepts a named list whose entries map column names to
either a single sentinel value or a vector of them
(`list(income = c(-9, -99, -999))`).  Only numeric columns are
affected; factor columns should be recoded with `NA` directly
before calling `robsynth()`.

## 12. Multiple synthesis with combining rules

Generating a single synthetic copy is enough for many exploratory
purposes, but for valid *inferential* statements the analyst needs
multiple copies plus pooling rules that account for synthesis
uncertainty.  Set `m > 1` in `robsynth()` and use `synth_lm()` /
`synth_glm()` to fit a model on each copy and return Reiter-pooled
estimates, standard errors, and degrees of freedom.

```{r multi}
res5 <- robsynth(df[, 1:6], m = 5, seed = 1)
fit  <- synth_lm(X4 ~ X1 + X2 + X3, res5)
pooled <- data.frame(
  term = names(fit$coefficients),
  coef = round(fit$coefficients, 4),
  se   = round(fit$se,           4),
  df   = round(fit$df,           1)
)
knitr::kable(pooled, row.names = FALSE,
             caption = "Table 2: Reiter (2003) partially-synthetic combined estimates (m = 5).")
```

Interpretation:

- `coef` is the mean of the five per-copy OLS estimates.  If the
  synthesiser recovers the true structure, these should be close to
  the true $(0.5, 0.5, 0.3)$ for $(X_1, X_2, X_3)$.
- `se` is $\sqrt{T_p} = \sqrt{\bar U + B/m}$.  The $\bar U$ part is
  what you would see on a single copy; $B/m$ adds the synthesis
  uncertainty.  With $m = 5$ copies, this extra contribution is
  modest when the synthesiser is stable.
- `df` comes from $\nu = (m-1)(1 + m \bar U / B)^2$ and can be very
  large when $\bar U$ dominates (confidence intervals then use the
  normal quantiles in the limit).

How many copies do you need?  A useful heuristic: $m = 5$ is enough
for publication-quality estimates; $m = 10$ reduces the between-copy
noise perceptibly; larger $m$ gives diminishing returns.  Reiter's
total variance $T_p = \bar U + B/m$ reduces to $\bar U$ as $B \to 0$;
with a single synthetic copy ($m = 1$) the between term vanishes
exactly, `df` becomes infinite, and the standard errors reduce to
those of an ordinary single-dataset regression.

Set `n_cores > 1` (Section 19) to parallelise across copies when `m`
is moderate-to-large; the per-draw seed machinery guarantees that
`seed = k` with `n_cores = 4` produces identical output to `seed = k`
with `n_cores = 1`.

## 13. Utility assessment

> **Important — the contamination paradox.** `utility()` compares
> the synthetic data against its stored `$original`, which *is* the
> **contaminated** training data.  A successful robust synthesiser
> produces synthetic data closer to the **clean** distribution than
> the training data was, so against the contaminated baseline it can
> look "worse" — its regression coefficients are in the right place,
> but the contaminated OLS baseline's coefficients are not, so the
> CIs do not overlap.  Always compute utility against a clean
> reference when one is available.

```{r utility_bad}
# Match column sets: res was trained before `income` was added in Section 8
cols <- names(res$synth)
u_vs_contam <- utility(res$synth,  df[, cols])                # contaminated
u_vs_clean  <- utility(res$synth,  df_clean[, cols])          # clean truth
data.frame(
  baseline       = c("contaminated training", "clean truth"),
  pMSE           = c(u_vs_contam$pMSE, u_vs_clean$pMSE),
  reg_CI_overlap = c(mean(u_vs_contam$regression$ci_overlap, na.rm = TRUE),
                     mean(u_vs_clean$regression$ci_overlap,  na.rm = TRUE)),
  reg_rel_bias   = c(u_vs_contam$regression$relative_bias,
                     u_vs_clean$regression$relative_bias)
) |>
  knitr::kable(digits = 3,
               caption = "Table 3: Utility against contaminated training data vs clean truth. The CI-overlap collapses to zero when measured against contamination because the synthesised coefficients sit near the TRUE values while OLS on the contaminated data does not.")
```

The pMSE is reported on the **raw** scale.  Under the null of perfect
synthesis the expected value is $c(1-c)/n$ with
$c = n_\mathrm{synth}/(n_\mathrm{orig} + n_\mathrm{synth})$ and $n$ the
total sample size; for an equal-size synthetic copy the null is
$1/(4n)$.  For cross-sample-size comparison, @snoke2018 recommend the
null-scaled $\mathrm{pMSE} / \mathrm{Var}(\mathrm{pMSE} \mid H_0)$.
Values near 1 are indistinguishable from a perfect synthesiser; larger
values indicate the classifier can tell original and synthetic records
apart.

Heuristic (not prescriptive) diagnostic guides for clean-baseline
utility:

- **Mean KS** < 0.05: marginals broadly preserved.
- **Correlation-Frobenius** < 0.1 (divided by the number of
  continuous variables): joint linear structure broadly preserved.
- **CI overlap** near 1: inferential conclusions broadly preserved.

Compared against a contaminated baseline, all three can look much
worse even when the synthetic data is objectively closer to the
truth.  The practical recipe:

1. If a clean reference is available (e.g. pre-contamination data,
   a validation pilot, a well-curated subset), run
   `utility(res$synth, clean_data)` to measure how well the
   synthesiser recovers the target distribution.
2. If only the contaminated training data is available, treat the
   utility metrics as diagnostic but not absolute.  Numbers that
   look "bad" may simply reflect that the synthesiser did the
   right thing and left the contamination behind.
3. Always pair utility with a disclosure-risk check (Section 16).
   A high-utility synthetic dataset that matches the original record
   for record is useless — it is the original dataset.

## 14. Model-based comparison

`utility()` produces aggregate metrics that summarise the synthetic
data as a whole.  Often the analyst cares about a *specific*
downstream model: "does my favourite regression on the synthetic
data give the same coefficients as on the original?"
`compare_fit()` answers exactly that.  Pass any formula, and the
function fits it on both the original and the synthetic data and
reports per-coefficient differences, standardised differences,
and confidence-interval overlaps.

```{r compare_fit}
comp <- compare_fit(X1 ~ X2 + X3 + G1, res, df, plot = FALSE)
print(comp)
```

The two headline numbers:

- **Mean CI overlap**: for each coefficient, the overlap of the
  original and synthetic 95 % CIs is computed and the mean is
  reported.  Values near 1 indicate near-perfect preservation;
  values near 0 indicate that the synthetic analysis would lead to
  different inferences.  @reiter2005 recommends > 0.5 as a
  practical threshold for "acceptable preservation".
- **Mean |standardised difference|**: each coefficient's difference
  divided by its pooled standard error, averaged across
  coefficients.  Values below 2 mean the shifts are within typical
  sampling noise.

Both numbers are affected by the contamination paradox discussed in
Section 13: if you compare against the contaminated original, a
successful robust synthesiser looks worse than a non-robust one.
For robust synthesis, pair `compare_fit()` with a clean reference
when you have one.

## 15. Visual comparison

Aggregate numbers are necessary but never sufficient — a good
synthesiser can have excellent pMSE and still miss the tails of a
bimodal distribution or get the sign of a correlation wrong.  The
visual diagnostics below cover three complementary perspectives:
univariate marginals, bivariate correlation structure, and the
pairwise joint distribution.  All three use the Okabe-Ito palette
for colourblind safety (original = black / grey, synthetic = orange).

**Figure 1 — marginals.**  `compare()` draws overlaid densities for
continuous variables and grouped bar plots for factors.  Each
continuous facet label shows the Kolmogorov-Smirnov statistic — a
crude but useful scalar summary of how close the two marginals are.
What to look for: matched modes, matched tails, no systematic shift.

```{r compare_marginals, fig.height = 5, fig.cap = "Figure 1: Marginal distributions and category proportions — original (black) vs synthetic (orange). KS statistic shown in each facet label.", fig.alt = "Density plots for continuous variables and bar plots for categorical variables, comparing original and synthetic data side by side."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  compare(res, df)
}
```

**Figure 2 — correlation structure, and a paradox.**  The natural
default is to compare two heatmaps: classical Pearson correlation on
the original data, classical Pearson correlation on the synthetic.
But under contamination **the classical original is itself biased**
— Pearson correlation has breakdown point zero, so a 15 % block
shift can drive its off-diagonal entries to anywhere.  If the robust
synthesiser recovered the *clean* correlation structure, its
classical panel will *differ* from the classical-original panel by
design.  Reading the two-panel comparison as "they should agree"
punishes successful robust synthesis — the same paradox as the
CI-overlap collapse in Section 13.

The 2×2 grid below lets the reader separate two independent
questions.  Rows = **how is correlation measured** (top = classical
Pearson, bottom = OGK-based robust, @maronna2019); columns =
**which dataset** (left = original, right = synthetic).

Reading the grid:

- **Top row** shows what the classical analyst would see.  Top-left
  is biased by the contamination; top-right is the synthetic Pearson
  correlation — which can match either the contaminated or the
  clean truth depending on whether the synthesiser left the
  contamination in.
- **Bottom row** is our best guess at what the **clean** correlation
  would look like.  Bottom-left and bottom-right should agree with
  each other *and* with the top-right if the synthesiser did its
  job; only the top-left is meant to be discrepant.
- **Columns** (original vs synthetic within each row) show how much
  the dataset-construction step changed the correlation when
  measured consistently.

Pass `include_robust_original = TRUE` to
`plot(..., type = "correlation")` to get this 2×2 rendering in your
own diagnostics; or pass `cor_method = "robust"` alone to get a
two-panel robust-vs-robust comparison.

```{r compare_correlation, fig.height = 5.4, fig.cap = "Figure 2: Correlation structure — 2x2 grid of heatmaps. Rows: classical Pearson (top) vs OGK-robust (bottom). Columns: original (left) vs synthetic (right). The top-left panel is biased by contamination; the remaining three panels should agree when the robust synthesiser recovered the clean structure.", fig.alt = "Four correlation heatmaps arranged in a 2x2 grid: rows are Pearson vs robust correlation, columns are original vs synthetic data."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  plot(res, type = "correlation", original = df,
       include_robust_original = TRUE)
}
```

**Figure 3 — pairwise joint distribution.**  Scatter plots of the
first three continuous variables reveal whether the synthesiser
captures the *joint* structure, not just the marginals.  Two
variables can have identical marginals and very different joint
behaviour (correlation, tail dependence, holes in the support), so
this plot is diagnostic in a way Figure 1 cannot be.

Pairwise scatter comes in two layouts: `"overlay"` (default) puts
both sources on a shared axis with transparency, and `"facet"` places
them in two side-by-side panels.  Overlay is faster to read when the
two clouds differ in *location*: you immediately see the orange
synthetic points sitting where the grey original points do (or do
not).  Facet is clearer when the clouds overlap heavily and only
their *shape* differs — you can see holes and density asymmetries
that the overlay smooths over.  The `alpha` argument controls point
transparency; its default (`0.4` for overlay, `0.6` for facet) is
usually fine.

```{r compare_scatter_overlay, fig.height = 4.5, fig.cap = "Figure 3a: Pairwise scatter, overlay layout — original (grey) and synthetic (orange) on a shared axis; alpha = 0.4.", fig.alt = "Scatter-plot matrix of the first three continuous variables with original and synthetic points overlaid and transparency applied."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  plot(res, type = "scatter", original = df, layout = "overlay")
}
```

```{r compare_scatter_facet, fig.height = 4.5, fig.cap = "Figure 3b: Pairwise scatter, facet layout — original (left panel) and synthetic (right panel) in separate sub-plots; alpha = 0.6.", fig.alt = "Scatter-plot matrix of the first three continuous variables with original and synthetic shown in side-by-side facets."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  plot(res, type = "scatter", original = df, layout = "facet")
}
```

## 16. Disclosure risk

A synthesiser that simply returned the training data would score
perfectly on every utility metric — and would leak every record.
Utility and disclosure risk are two sides of the same coin, and any
release workflow must track both.  Risk is harder to measure than
utility because it depends on the *attacker*: what side information
do they have, which records are they targeting, what model do they
use?  `disclosure()` wraps several complementary measures so the
analyst gets a multi-model picture rather than betting on a single
number.

The built-in `"distance"` measure is a cheap first check that works
on numeric keys only: it scales each continuous key by its MAD and
reports the distribution of nearest-neighbour distances between
original and synthetic records.  When the **riskutility** package is
installed, six additional model-based measures become available,
each capturing a different attacker model: `"rapid"` [@thees2025] (a
random forest attacks the sensitive attribute), `"dcap"`
(attribution via exact or Gower matching), `"dcr"`
(distance-to-closest-record with a holdout null distribution),
`"delta_presence"` (membership inference via $\delta$-presence),
`"domias"` (density-based membership inference), and `"drisk"`
(sdcMicro-style risk).  Pass a character vector of method names or
the shorthand `"all"` to compute every available measure in one
call.

```{r disclosure_basic}
d_basic <- disclosure(res, df,
                      key_vars   = c("X1", "X3"),
                      target_var = "X2",
                      method     = "distance")
print(d_basic)
```

The printed statistics are **descriptive**, not a prescription:

- **median nearest** close to 0 means many synthetic records sit on
  top of originals in quasi-identifier space — a warning sign;
- **median nearest** much larger than the typical within-sample
  spacing means the synthesiser moved records far from their
  training counterparts — usually a good sign for privacy, but check
  that utility has not collapsed;
- **q05** and **% below 0.5** are tail-sensitive summaries that
  capture whether the *closest* synthetic records are especially
  close — a small number of leaked records can dominate the risk
  even when the median looks healthy.

There is no universal "good" threshold; values depend on sample
size, the number of quasi-identifiers, and how spread out the
training data already are.  Compare against a clearly-privacy-safe
baseline such as a pure marginal resample to calibrate what "far"
means for your data.

### 16a. RAPID (model-based attacker)

RAPID trains a random-forest attacker on the synthetic data and scores
how well it can predict the sensitive attribute on the originals.
Unlike distance-based risk, it handles mixed-type keys natively and
gives a single, directly comparable [0, 1] score.

```{r disclosure_rapid, eval = requireNamespace("riskutility", quietly = TRUE)}
d_rapid <- disclosure(res, df,
                      key_vars   = c("X1", "X3", "G1", "G2"),
                      target_var = "X2",
                      method     = "rapid")
d_rapid$rapid
```

### 16b. Multiple measures in one call

When several measures are requested, `disclosure()` runs each in turn
and collects the results.  The example below exercises four
complementary perspectives:

```{r disclosure_all, eval = requireNamespace("riskutility", quietly = TRUE)}
d_all <- disclosure(
  res, df,
  key_vars   = c("X1", "X3", "G1", "G2"),
  target_var = "X2",
  method     = c("rapid", "dcap", "domias", "delta_presence"))
# access individual measures via the list components:
names(d_all)
d_all$dcap
```

The qualitative picture reported by these measures is usually
consistent; disagreements between RAPID (model-based) and DCR
(density-based) often flag issues with one of the attacker models
rather than true differences in disclosure risk — for example, RAPID
will show elevated scores when the synthesiser preserved a strong
predictive relationship that a real attacker could also exploit, and
DCR will show elevated scores when individual records are too close
in Gower / Euclidean space.  Triangulating across measures is safer
than trusting any single one.

A typical release decision flow:

1. Check utility (Section 13) against the cleanest available
   reference.  If utility is insufficient, no release.
2. Compute `disclosure(..., method = "all")`.  If any measure flags
   catastrophic risk (RAPID above 0.15, median DCR below the
   privacy-safe baseline, DOMIAS AUC near 1), investigate and either
   fix the synthesiser or withhold the release.
3. Publish the synthetic copy together with the disclosure numbers
   and the exact command that produced it (for reproducibility).

## 17. Hierarchical data (household / person)

Household surveys, education cohorts, longitudinal panels — many
real datasets come in two-level form where records within a cluster
are not exchangeable with records in other clusters.  Naively
synthesising such data at the record level destroys the cluster
structure (e.g. producing "households" whose members' incomes
contradict the household total).  `robsynth_hh()` handles this with
a two-phase procedure:

- **Phase 1** resamples whole households with replacement.  Each
  synthetic household keeps its member-count and its
  household-level variables (income, region, size) intact — these
  are *not* synthesised, they are bootstrapped.
- **Phase 2** synthesises the person-level variables (age, sex,
  etc.) conditional on the household-level variables and previously
  synthesised person variables.  The conditional models are fit on
  the **original** full data and predicted on the resampled
  household structure, which means the person-level distributions
  reflect the full empirical population, not just the bootstrapped
  households.

```{r hierarchical}
set.seed(1)
hh <- data.frame(
  hhid      = rep(1:100, each = 3),
  hh_income = rep(rnorm(100, 5000, 1000), each = 3),
  region    = factor(rep(sample(c("A", "B"), 100, TRUE), each = 3)),
  age       = rnorm(300, 40, 15),
  sex       = factor(sample(c("M", "F"), 300, TRUE))
)
hh$hh_income[1:15] <- hh$hh_income[1:15] + 8000            # hh outliers

res_hh <- suppressMessages(
  robsynth_hh(hh, hhid = "hhid",
              hh_vars     = c("hh_income", "region"),
              person_vars = c("age", "sex"),
              seed = 1)
)
print(res_hh)
all(table(res_hh$synth$hhid) == 3)                         # sizes preserved
```

## 18. Survey weights

Most official-statistics microdata carry **design weights** that
reflect the sampling scheme: stratified, multi-stage, post-stratified,
raked.  Ignoring these weights during synthesis can bias conditional
models because units with small probability of selection are
over-represented in the unweighted fit.

`robsynth()` accepts weights via the `weights =` argument.  Pass
either a variable name (the column is removed from the synthesised
variables and used only as weights) or a numeric vector of length
`nrow(data)`.  Weights are propagated to every conditional fit: MM,
weighted logistic, robust tree-based methods — all of them accept
case weights and use them.

```{r weights}
set.seed(456)
df_w <- df[, 1:6]
df_w$sampweight <- runif(nrow(df_w), 1, 10)

res_w <- robsynth(df_w, weights = "sampweight", seed = 1)
names(res_w$synth)                                         # weight not synthesised
```

Point estimates are approximately design-consistent.  The standard
errors reported by `lmrob` are **model-based**, not design-corrected;
for design-based inference wrap the synthesiser in a replicate-weights
or bootstrap procedure [@lumley2010].

## 19. Parallel computation

When `m` is large (50 bootstrap replications for a simulation study;
20 synthetic copies for publication-quality Reiter inference), the
synthesis loop dominates the runtime.  Set `n_cores > 1` to
distribute the `m` independent synthetic copies across CPU cores via
`parallel::parLapply`.

```{r parallel, eval = FALSE}
res_par <- robsynth(df, m = 20, n_cores = 4, seed = 1)
```

Per-draw seeds (`seed + 1:m`) are set inside each worker before any
random draw, so the parallel and sequential paths produce **identical**
output under a fixed `seed`.  This is not the case for many other
parallel R packages; if you rely on byte-for-byte reproducibility,
verify it for your specific combination of BLAS, OS, and R version.

Practical recommendations:

- Start with `n_cores = parallel::detectCores() - 1` to leave one
  core free for the OS.
- The parallel overhead is ~1 second per worker, so parallelisation
  pays off only for `m` in the tens with non-trivial per-copy cost.
- `n_cores > 1` is ignored when `m = 1`.

## 20. Diagnosing convergence or fit failure

> **Warning.** A `NULL` entry in `res$models` signals a silent fit
> failure: the continuous path fell back to `lm()` or the categorical
> path fell back to marginal sampling.  Always inspect
> `sapply(res$models, is.null)` before release, and re-run with an
> alternative method for any variable whose model is `NULL`.

Occasionally `lmrob` fails to converge (severe singular design,
fewer rows than predictors after NA omission, or a factor with a
singleton level after contamination).  The default continuous path
silently falls back to `lm()` in that case; for categorical paths the
stored model is `NULL` and the generator falls back to marginal
sampling — i.e. the variable is drawn from its empirical marginal
distribution, ignoring all conditioning.

Silent fallback is deliberate: a hard error in the middle of a
sequential synthesis would leave you with a partially-populated
synthetic dataset.  But it does mean you need to **check**.

Practical checks:

```{r diag_models, eval = FALSE}
# after `res <- robsynth(df, ...)`:
sapply(res$models, is.null)           # NULL marks silently-failed variables
```

Remedies for the most common failure modes:

- **Factor with a singleton class after contamination** — collapse
  rare classes before synthesis (`droplevels()` followed by
  `forcats::fct_lump_min(min = 2)`), or explicitly use
  `method = "sample"` for that variable.
- **`robust_xgboost` with unseen factor levels on new data** —
  apply `droplevels()` to the training data first so the training
  `xlev` is exhaustive.
- **Multicollinearity or singular design** — inspect the predictor
  matrix (Section 10); drop one of the collinear predictors from
  the row of the problematic variable.
- **Very small sample** ($n < 2p$) — try `method = "cart"` or
  `method = "sample"` for the affected variables; MM needs enough
  degrees of freedom to estimate its scale.

## 21. Migration from synthpop

If you already have a **synthpop** workflow, migrating to **robSynth**
is mechanical: the argument names and calling conventions are
deliberately compatible.  The table below lists the feature mapping.

| Feature | **synthpop** | **robSynth** |
|---------|--------------|--------------|
| Default method | CART | MM-estimation |
| Robustness | None | Bounded influence |
| Per-variable methods | Yes | Yes |
| Custom formulas | Via `passive` | Direct `formulas =` |
| Multiple synthesis | Yes | Yes (`synth_lm()` / `synth_glm()`) |
| Hierarchical data | Limited (`nested`) | Yes (`robsynth_hh()`) |
| Survey weights | No | Yes |
| Disclosure risk | CAP / TCAP | Six `riskutility` measures (RAPID, DCAP, DCR, $\delta$-presence, DOMIAS, drisk) + built-in distance |
| CRAN | Yes | Planned |

```{r migration, eval = FALSE}
# synthpop:
library(synthpop)
syn(data, method = "cart", m = 5)

# robSynth equivalent (non-robust, same behaviour):
robsynth(data, method = "cart", m = 5)

# robSynth default (robust):
robsynth(data, m = 5)
```

The main behavioural difference you will notice is in the default:
**synthpop**'s default CART is replaced by **robSynth**'s MM.  On
clean data the two produce comparable synthetic copies; on
contaminated data MM preserves the joint structure while CART
partitions on the outliers.  If you want the synthpop default
behaviour verbatim, pass `method = "cart"`.

## 22. Summary

**robSynth** plugs into the sequential-synthesis idiom of **synthpop**
and **simPop** but replaces every OLS/CART conditional with a
bounded-influence alternative.  The headline guarantees — 50 %
breakdown for MM, conditional y-outlier flagging via OOB residuals for
`robust_rf`, pseudo-Huber loss for `robust_xgboost`, Reiter's partial
combining for $m>1$ inference — are summarised in Section 2 and
demonstrated in Sections 4 onward.  The package is designed for
practitioners: per-variable methods, sentinel-code recoding, survey
weights, hierarchical data, and six **riskutility** disclosure-risk
measures (RAPID, DCAP, DCR, $\delta$-presence, DOMIAS, drisk) are
first-class citizens.  Section 13 highlights the "contamination
paradox": utility metrics computed against the contaminated training
data penalise successful robust synthesis.  A simulation study
accompanying the paper [Templ 2026, submitted] quantifies the
population-level story for both block and cellwise contamination;
the ranking of methods in any single demo should be interpreted as
illustrative rather than definitive.

## References

<div id="refs"></div>

## Session info

```{r sessioninfo}
sessionInfo()
```
