---
title: "robSynth: Robust Synthetic Data Generation"
author: "Matthias Templ"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{robSynth: Robust Synthetic Data Generation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  message = FALSE,
  warning = FALSE
)
```

## Installation

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("statistikat/robSynth")
```

## The problem: contamination propagates through synthesis

Standard synthetic data generators (synthpop, simPop) use OLS or CART to
model each variable sequentially.  When the training data contains
outliers, each conditional model is distorted, and the distortion
compounds through subsequent variables.

**robSynth** replaces these non-robust models with MM-estimators
(continuous) and Huber-weighted logistic regression (categorical).
MM-estimation achieves 50% breakdown point (the maximum for regression)
and 95% asymptotic efficiency under Gaussian errors.

> **When to use robSynth instead of synthpop:** whenever your training data
> may contain outliers, coding errors, linkage mismatches, or measurement
> artifacts.  If you are certain the data is clean, synthpop's CART is a
> fine choice.

## The key demo: robust vs non-robust under contamination

```{r demo}
library(robSynth)

# Generate data with a known regression structure
set.seed(42)
n <- 500
df <- data.frame(
  X1 = rnorm(n),
  X2 = rnorm(n),
  X3 = rnorm(n),
  X4 = rnorm(n)
)
# True relationship: X4 = 0.5*X1 + 0.5*X2 + 0.3*X3 + noise
df$X4 <- 0.5 * df$X1 + 0.5 * df$X2 + 0.3 * df$X3 + rnorm(n, sd = 0.5)
df$G1 <- factor(rbinom(n, 1, plogis(0.3 * df$X1)))
df$G2 <- factor(rbinom(n, 1, plogis(0.3 * df$X2)))

# Add 15% contamination: large shift in X1 and X2
idx <- sample(n, 75)
df$X1[idx] <- df$X1[idx] + 8
df$X2[idx] <- df$X2[idx] + 8
df$G1[idx] <- factor(1 - as.integer(df$G1[idx]) + 1)

# Store clean version
df_clean <- df
df_clean$X1[idx] <- df_clean$X1[idx] - 8
df_clean$X2[idx] <- df_clean$X2[idx] - 8

# Compare: robust vs non-robust synthesis (10 reps averaged)
true_coef <- c(X1 = 0.5, X2 = 0.5, X3 = 0.3)  # true betas for X4 model
rmse_mm <- rmse_ols <- cor_mm <- cor_ols <- numeric(10)
for (r in 1:10) {
  s_mm  <- robsynth(df, seed = r)$synth
  s_ols <- robsynth(df, method = "parametric", seed = r)$synth
  # Regression on synthetic data vs true coefficients
  fit_mm  <- coef(lm(X4 ~ X1 + X2 + X3, data = s_mm))[c("X1","X2","X3")]
  fit_ols <- coef(lm(X4 ~ X1 + X2 + X3, data = s_ols))[c("X1","X2","X3")]
  rmse_mm[r]  <- sqrt(mean((fit_mm - true_coef)^2))
  rmse_ols[r] <- sqrt(mean((fit_ols - true_coef)^2))
  # Correlation preservation
  cor_mm[r]  <- sqrt(sum((cor(df_clean[,1:4]) - cor(s_mm[,1:4]))^2))
  cor_ols[r] <- sqrt(sum((cor(df_clean[,1:4]) - cor(s_ols[,1:4]))^2))
}

cat("                RMSE-beta  Corr Frobenius\n")
cat(sprintf("MM (robust):    %.4f     %.4f\n", mean(rmse_mm), mean(cor_mm)))
cat(sprintf("OLS:            %.4f     %.4f\n", mean(rmse_ols), mean(cor_ols)))
cat(sprintf("MM advantage:   %.0f%%        %.0f%%\n",
            100 * (1 - mean(rmse_mm)/mean(rmse_ols)),
            100 * (1 - mean(cor_mm)/mean(cor_ols))))
```

Under 15% contamination with large leverage shifts (+8 SD in X1 and
X2), MM synthesis preserves both regression coefficients and correlation
structure significantly better than OLS.  The contaminated observations
pull OLS regression slopes toward zero (attenuation bias), while MM
downweights them and recovers the true conditional relationships.

## Basic usage

```{r basic}
# Default: robust synthesis (MM + weighted logistic)
res <- robsynth(df, seed = 1)
print(res)
```

The result stores both the synthetic data (`res$synth`) and the original
(`res$original`) for convenient downstream analysis.

## Per-variable methods

Mix robust and non-robust methods as needed:

```{r pervar}
res2 <- robsynth(df,
  method = c(X1 = "mm", X2 = "robust_rf", X3 = "robust_xgboost",
             X4 = "mm", G1 = "robust_logreg", G2 = "polyreg"),
  seed = 1)
res2$method
```

Available methods:

| Method | Type | Robustness |
|--------|------|-----------|
| `"mm"` | Continuous (default) | MM-estimation, 50% breakdown |
| `"robust_cart"` | Both | Median leaves + MAD scale |
| `"robust_rf"` | Both | Weighted by y-outlier distance |
| `"robust_xgboost"` | Both | Pseudo-Huber loss |
| `"robust_logreg"` | Categorical (default) | Huber-weighted logistic |
| `"norm"` | Continuous | Standard OLS (no robustness) |
| `"cart"` / `"ctree"` | Both | Standard trees |
| `"logreg"` / `"polyreg"` | Categorical | Standard logistic/multinomial |
| `"sample"` | Both | Marginal resampling |
| `"passive"` | Derived | Deterministic formula |

**Note:** MM-estimation assumes approximately symmetric conditional
errors.  For heavily skewed variables, apply a transformation first
(see next section).

## Transformations for skewed variables

```{r transform}
# Add an income-like variable (right-skewed) for this demo
df$income <- rlnorm(n, 10, 0.5)
df$income[idx] <- df$income[idx] * 10

# Robust Box-Cox normalizes before synthesis, back-transforms after
res3 <- robsynth(df, transform = list(income = "boxcox"), seed = 1)
cat("Estimated lambda:", res3$transform_info$income$lambda, "\n")
# Lambda near 0 = log-transform (expected for lognormal data)
```

The lambda parameter is estimated by minimizing a robust skewness
criterion (`|mean - median| / MAD`), which is itself resistant to the
outliers that motivate the transformation.

## Custom formulas and visit sequence

```{r formulas}
res4 <- robsynth(df,
  visit.sequence = c("G1", "G2", "X1", "X2", "X3", "X4", "income"),
  formulas = list(X3 = ~ X1 * X2 + G1),
  transform = list(income = "log"),
  seed = 1)
```

## Multiple synthesis with combining rules

Generate `m > 1` synthetic copies for valid statistical inference:

```{r multi}
res5 <- robsynth(df[, 1:6], m = 5, seed = 1)

# Fit a model on the synthetic data with Reiter (2003) combining rules
# (NOT Rubin's rules -- the variance formula differs for synthetic data)
fit <- synth_lm(X1 ~ X2 + X3 + G1, res5)
data.frame(
  coef = round(fit$coefficients, 4),
  se   = round(fit$se, 4),
  df   = round(fit$df, 1)
)
```

## Utility assessment

```{r utility}
u <- utility(res)
print(u)
```

Interpretation:

- **pMSE** near 0: synthetic data is hard to distinguish from original
- **Mean KS** < 0.05: marginal distributions well preserved
- **Correlation** < 0.1: joint structure well preserved
- **CI overlap** near 1: inferential conclusions preserved

## Model-based comparison

Compare a specific regression analysis on original vs synthetic:

```{r compare_fit}
comp <- compare_fit(X1 ~ X2 + X3 + G1, res, df, plot = FALSE)
print(comp)
```

A mean CI overlap above 0.5 and mean |standardized difference| below 2
indicate acceptable preservation for most applications.

## Visual comparison

```{r compare_plot, fig.height = 5}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  compare(res, df)
}
```

## Disclosure risk

Assess whether a random-forest attacker could infer sensitive attributes
from the synthetic data:

```{r disclosure}
d <- disclosure(res, df,
  key_vars = c("X1", "G1"),
  target_var = "X2",
  method = "distance")
print(d)
```

Interpretation of distance-based risk:

- **Median distance** > 1.0: low risk (synthetic records are far from
  originals)
- **5th percentile** > 0.3: even the closest matches are reasonably
  distant
- **% below threshold** < 10%: few records at high disclosure risk

When the `riskutility` package is installed, RAPID (Thees, Mueller &
Templ, 2025) provides a model-based risk measure using the proportion of
records whose sensitive attribute can be predicted within 10% error.

## Hierarchical data (household/person)

For survey data with nested structures:

```{r hierarchical}
set.seed(1)
hh <- data.frame(
  hhid      = rep(1:100, each = 3),
  hh_income = rep(rnorm(100, 5000, 1000), each = 3),
  region    = factor(rep(sample(c("A", "B"), 100, TRUE), each = 3)),
  age       = rnorm(300, 40, 15),
  sex       = factor(sample(c("M", "F"), 300, TRUE))
)
# Add household-level outliers
hh$hh_income[1:15] <- hh$hh_income[1:15] + 8000

res_hh <- suppressMessages(
  robsynth_hh(hh, hhid = "hhid",
              hh_vars = c("hh_income", "region"),
              person_vars = c("age", "sex"),
              seed = 1)
)
print(res_hh)

# Verify: household sizes preserved
all(table(res_hh$synth$hhid) == 3)
```

Phase 1 resamples complete households (preserving within-household
composition), Phase 2 synthesizes person-level variables conditional on
household context.

## Survey weights

Design weights from complex sampling are passed to all model fits:

```{r weights}
df_w <- df[, 1:6]
df_w$sampweight <- runif(nrow(df_w), 1, 10)

res_w <- robsynth(df_w, weights = "sampweight", seed = 1)
# Weight column used for fitting but not synthesized
names(res_w$synth)
```

Note: survey weights modify the estimation objective.  Point estimates
are approximately design-consistent, but standard errors from the
weighted lmrob are not fully design-corrected.

## Parallel computation

When generating many synthetic copies, distribute across cores:

```{r parallel, eval = FALSE}
res_par <- robsynth(df, m = 20, n_cores = 4, seed = 1)
```

## Comparison with synthpop

| Feature | synthpop | robSynth |
|---------|----------|----------|
| Default method | CART | MM-estimation |
| Robustness | None | Bounded influence |
| Per-variable methods | Yes | Yes |
| Custom formulas | Via `passive` | Direct `formulas` argument |
| Multiple synthesis | Yes | Yes (+ `synth_lm()`) |
| Hierarchical data | Limited (`nested`) | Yes (`robsynth_hh()`) |
| Survey weights | No | Yes |
| Disclosure risk | Yes (CAP/TCAP) | Yes (RAPID) |
| CRAN | Yes | Planned |

**Migration from synthpop:**

```{r migration, eval = FALSE}
# synthpop:
library(synthpop)
syn(data, method = "cart", m = 5)

# robSynth equivalent (non-robust, same behavior):
robsynth(data, method = "cart", m = 5)

# robSynth default (robust):
robsynth(data, m = 5)
```

## Citation

```
Templ, M. (2026). Robust synthetic data generation: protecting data
quality from contaminated training data. Submitted.
```
